---
name: Cyclistic Bike Share Company
surname: 
position: "A case study by Pierpaolo Zotti"
address: "Ilha de Guaratiba - Rio de Janeiro"
phone: +55 21 98130 2437
www: pierpaolozotti.com
email: "zotti.pierpaolo@outlook.it"
twitter: 
github: "PierpaoloZotti"
linkedin: "pierpaolo-zotti-186399253"
date: "`r format(Sys.time(), '%B %Y')`"
output: 
  vitae::awesomecv:
    page_total: true
---
```{r Settings, message=FALSE, warning=FALSE, include=FALSE}
options(repos = list(CRAN="http://cran.rstudio.com/"))
Sys.setlocale("LC_TIME", "C") #For data manipulation 
```

# Introduction

Welcome to my personal analysis of the ***Cyclistic Bike Share Company***.
In order to answer the *key business questions*, I will follow the steps of the data analysis process: **ask**, **prepare**, **process**, **analyze**,
**share**, and **act**.

# About the company

In 2016, **Cyclistic** launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that are geotracked and locked 
into a network of 692 stations across Chicago. The bikes can be unlocked from one station and returned to any other station in the system anytime.

Cyclistic’s finance analysts have concluded that annual members are much more profitable than casual riders. Although the pricing flexibility helps Cyclistic attract more customers, Moreno believes that maximizing the number of annual members will be key to future growth. Rather than creating a marketing campaign that targets all-new customers, is a very good chance to convert casual riders into members. 

# Company goal

The company final goal is to increase the number of annual member.
My task as junior analyst is presentig a clear visualization and some recommendations about the strategy to accomplish that goal.


# Data source

The data for this analysis is the Divvy dataset, and we will use the last
12 months to accomplish our scope.

# Installing required packages

For this case study we will use`tidyverse` for importing and wrangling data,
`lubridate` for data manipulation, and `ggplot2` for data visualizations.

```{R Packages Installations, echo=TRUE, message=FALSE, warning=FALSE}
install.packages("tidyverse")
install.packages("lubridate")
install.packages("ggplot2")
install.packages("scales")
library(tidyverse)
library(lubridate)
library(ggplot2)
library(scales)
```

# Check and set the working directoy

```{r Set working directory, echo=TRUE, message=FALSE, warning=FALSE}
getwd() #Display the actual working directory
setwd("/Users/zotti/OneDrive/Desktop/Case Study 1/csv")
```

# Collect data from csv files

```{r Data Upload, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
td2021_10 <- read_csv("csv/202110-divvy-tripdata.csv")
td2021_11 <- read_csv("csv/202111-divvy-tripdata.csv")
td2021_12 <- read_csv("csv/202112-divvy-tripdata.csv")
td2022_01 <- read_csv("csv/202201-divvy-tripdata.csv")
td2022_02 <- read_csv("csv/202202-divvy-tripdata.csv")
td2022_03 <- read_csv("csv/202203-divvy-tripdata.csv")
td2022_04 <- read_csv("csv/202204-divvy-tripdata.csv")
td2022_05 <- read_csv("csv/202205-divvy-tripdata.csv")
td2022_06 <- read_csv("csv/202206-divvy-tripdata.csv")
td2022_07 <- read_csv("csv/202207-divvy-tripdata.csv")
td2022_08 <- read_csv("csv/202208-divvy-tripdata.csv")
td2022_09 <- read_csv("csv/202209-divvy-publictripdata.csv")
```

# Check data consistency

Checking column names with`colnames()` function and the structure with `str()`
I notice that all the columns of each dataset has a consistent name and data 
type

# Join all dataset together 

```{r Join Dataframes, echo=TRUE, message=FALSE, warning=FALSE}
trips_df <- bind_rows(td2021_10,
                      td2021_11,
                      td2021_12,
                      td2022_01,
                      td2022_02,
                      td2022_03,
                      td2022_04,
                      td2022_05,
                      td2022_06,
                      td2022_07,
                      td2022_08,
                      td2022_09)
```

# Checking for NULL values

Using the summary() function on the dataset I notice that there are 5844
rows of NULL values only on the Lat and Lng Columns.This is only the 1%
of the total rows of our dataset. We can exclude these rows.
But looking at the table with th View() function I notice that there are
other N/A values on the start_station_name, end_statiom_name, start_station_id.

I will use the colSums(is.na()) function to count how many N/A values we have:

```{r Check NA values, echo=TRUE, message=FALSE, warning=FALSE}
colSums(is.na(trips_df))

```
This is 16% of total rows.

# Excluding unnecessary columns

For this case study we don't need the information start_station_id and end_station_id

```{r Excluding Unecessary Columns, echo=TRUE, message=FALSE, warning=FALSE}
trips_df <- trips_df %>% 
  select(-c(start_station_id, 
            end_station_id))
```

# Check member_casual and rideable_type field for unwanted records

```{r Variable Check for desired columns, echo=TRUE}
print("Bike type:")
table(trips_df$rideable_type)
print("User type:")
table(trips_df$member_casual)
```

# Add column date, year, month, day, and day_of_week

This is for aggregation purpose. By now only ride level aggregation is possible.

```{r Date Manipulation, echo=TRUE, message=FALSE, warning=FALSE}
trips_df$date <- as.Date(trips_df$started_at)
trips_df$year <- format(as.Date(trips_df$date),"%Y")
trips_df$month <- format(as.Date(trips_df$date),"%m")
trips_df$day <- format(as.Date(trips_df$date),"%d")
trips_df$day_of_week <- format(as.Date(trips_df$date),"%A")
```

# Add column ride_lengthusing difftime() function

```{r Insert ride_length field, echo=TRUE, message=FALSE, warning=FALSE}
trips_df$ride_length <- difftime(trips_df$ended_at, trips_df$started_at)
```

Now we assure that ride_length must be numeric

```{r Casting Column ride_length, echo=TRUE, message=FALSE, warning=FALSE}
trips_df$ride_length <- as.numeric(as.character(trips_df$ride_length))

```
# Clenaing data 

I noticed that some rows of started_at is greater or equals than ended_at, 
generating negative or zero values of ride_length (Inconsistent values).
In addition I will remove all the rides with more than 24 hours, cause
based on Divvy [site](https://help.divvybikes.com/hc/en-us/articles/360033484791-What-if-I-keep-a-bike-out-too-long-), after a 24 hours period the customer may be charged with a lost
or stolen bike fee of 1200$

```{r Clean Inconsistent Rows, echo=TRUE, message=FALSE, warning=FALSE}
trips_df <- trips_df[!(trips_df$ride_length<=1),]
trips_df <- trips_df[!(trips_df$ride_length>24*60*60),]

```

# Analysis

I will begin analyzing some values of trip duration for casual and member
user:

```{r Comparsion Values, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
aggregate(trips_df$ride_length ~ trips_df$member_casual, FUN = min) # Shortest ride
aggregate(trips_df$ride_length ~ trips_df$member_casual, FUN = mean) #Straight average
aggregate(trips_df$ride_length ~ trips_df$member_casual, FUN = median) # Midpoint
aggregate(trips_df$ride_length ~ trips_df$member_casual, FUN = max) # Longest ride
```

Now I will calculate the average ride_length for users by day_of_week.

```{r User_week_stats, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
trips_df$day_of_week <- ordered(trips_df$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

trips_df %>% 
  aggregate(ride_length ~ member_casual + day_of_week, FUN = mean)
```



```{r Visualizations, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
trips_df %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n(),average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, weekday) %>% 
  as.data.frame()

trips_df %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title = "Ridership by customer type and weekday", 
       caption = "from 2021-10 to 2022-09", x="WEEKDAY" , y="NUMBER OF RIDES")+
  theme(axis.text.y = element_text(angle = 45))+ 
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))

trips_df %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title = "Average trip duration by customer type and weekday", 
       caption = "from 2021-10 to 2022-09", x="WEEKDAY" , y="TRIP DURATION(secs)")+
  theme(axis.text.y = element_text(angle = 45))+ 
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))


trips_df %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,max_duration = max(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = max_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title = "Max trip duration by customer type and weekday", 
       caption = "from 2021-10 to 2022-09", x="WEEKDAY" , y="TRIP DURATION(secs)")+
  theme(axis.text.y = element_text(angle = 45))+ 
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))

```




# Exporting data for further analysis

```{r Export Data to CSV, echo=TRUE}
counts <- aggregate(trips_df$ride_length ~ trips_df$member_casual + 
                      trips_df$day_of_week, FUN = mean)
df_by_month <- trips_df %>% 
  group_by(month, member_casual) %>% 
  drop_na() %>% 
  summarize(avg_ride_length = mean(ride_length),
            max_ride_length = max(ride_length))

df_by_bike_user <- trips_df %>% 
  group_by(rideable_type, member_casual) %>% 
  summarize(number_of_ride = n(), max_duration = max(ride_length),
            avg_duration_secs = round(mean(ride_length), digits = 2))

df_by_lat_lng <- trips_df %>% 
  group_by(start_lat, start_lng) %>% 
  summarize(number_of_ride = n()) %>%
  filter(number_of_ride>50) %>% 
  arrange(-number_of_ride)

df_by_start_station <- trips_df %>% 
  group_by(start_station_name, end_station_name) %>% 
  summarize(number_of_ride = n()) %>%
  filter(number_of_ride>50) %>% 
  arrange(-number_of_ride)

# Create the roundtrip column

trips_df$roundtrip <- ifelse(trips_df$start_station_name == trips_df$end_station_name,1,0)

# Create a dataframe with sum of roundtrip grouped by station, usertype, month and day

df_roundtrip <- trips_df %>% 
  filter(roundtrip==1) %>% 
  group_by(start_station_name, member_casual, month, day_of_week) %>% 
  summarize(number_of_ride = n()) %>% 
  arrange(-number_of_ride)
# Create a df with rides longer than 45 min to study income from member's rides

df_renv_member<- trips_df[!(trips_df$ride_length < 45*60),]
df_renv_member <- df_renv_member %>% 
  filter(member_casual=="member")

# Create a df with rides longer than 180 min to study income from casual's rides

df_renv_casual<- trips_df[!(trips_df$ride_length < 180*60),]
df_renv_casual <- df_renv_casual %>% 
  filter(member_casual=="casual")

# Create a csv file to analize with other software

write.csv(counts, file = 'C:/Users/zotti/OneDrive/Desktop/Case Study 1/csv/avg_ride_lenght.csv')
write.csv(df_by_month, file= 'C:/Users/zotti/OneDrive/Desktop/Case Study 1/csv/df_by_month.csv')
write.csv(df_by_bike_user, file= 'C:/Users/zotti/OneDrive/Desktop/Case Study 1/csv/df_by_bike.csv')
write.csv(df_by_start_station, file= 'C:/Users/zotti/OneDrive/Desktop/Case Study 1/csv/df_by_station.csv')
write.csv(df_by_lat_lng, file= 'C:/Users/zotti/OneDrive/Desktop/Case Study 1/csv/df_by_latlng.csv')
write.csv(df_roundtrip, file= 'C:/Users/zotti/OneDrive/Desktop/Case Study 1/csv/df_roundtrip.csv')
write.csv(df_renv_casual, file= 'C:/Users/zotti/OneDrive/Desktop/Case Study 1/csv/df_renv_casual.csv')
write.csv(df_renv_member, file= 'C:/Users/zotti/OneDrive/Desktop/Case Study 1/csv/df_renv_member.csv')
```
```

